---
title: "submission1"
author: "Antara Sengupta"
date: "2024-07-27"
output: html_document
---

## Loading in necessary packages
```{r loading packages}
library(readr)
library(dplyr)
library(ggplot2)
library(tibble)
library(tidyr)
library(dplyr)
```

## Loading Data

```{r loading data}
#loading in the two separate datasets provided to us (metadata & gene exppression data)
series <- read.csv("data/QBS103_GSE157103_series_matrix.csv")
genes <- read.csv("data/QBS103_GSE157103_genes.csv")

#displaying first few rows of each dataframe to get familiarized with the data
head(series)
head(genes)
```
## Formatting and cleaning data

```{r adjusting dataframes}
#goal is to merge both dataframes - we can merge them based on participant ID

#transposing the genes df and turning it back to a df so it can have 'participant_id' as rows (the same rows as the metadata)
genes_transposed <- as.data.frame(t(genes))

# getting rid of the x row because it has redundant info (same values as column names)
genes_transposed <- genes_transposed[-1, ]

#setting the row name to participant id so we can use this column to merge with metadata
genes_transposed <- rownames_to_column(genes_transposed, var = "participant_id")

#reshaping the gene so that genes and expression can be their own columns using pivot_longer
#used this source to learn how to implement pivot_longer: https://tidyr.tidyverse.org/reference/pivot_longer.html
genes_long <- genes_transposed %>%
   pivot_longer(
     cols = -participant_id,  #accessing all columns with the exception of participant_id because it is the only column without gene info
     names_to = "gene", #setting name column to gene
     values_to = "expression"  #setting values column to expression
   )

#merging long gene data and series metadata to become one dataframe
all_data <- merge(genes_long, series, by = "participant_id") #merging it on 'participant_id' column

#taking a look at the new comprehensive dataframe
head(all_data)
```


## Choosing focuses of interests: genes, continuous covariates and categorical covariates
#- gene: A1BG
#- continuous covariate: ferritin
#- categorical covariate: disease status, gender or icu status 
```{r subsetting dataframe}
#subsetting the dataframe so that it only contains columns that I want to perform further analysis on 
#using select to subset https://www.educative.io/answers/what-is-the-select-function-in-r
covid_data <- all_data %>% select(participant_id, gene,expression, ferritin.ng.ml., sex, disease_status)

#taking a look at the new subsetted data frame
head(covid_data)
```

## Plotting gene expression (histogram)
```{r histogram}
# Filter for gene A1BG and select expression column
a1bg_data <- df %>%
  filter(gene == "A1BG") %>%
  select(expression) %>%
  mutate(expression = as.numeric(expression))  # Ensure expression is numeric

# Plot histogram
ggplot(a1bg_data, aes(x = expression)) +
  geom_histogram(binwidth = 0.5, color = "black", fill = "blue") +
  labs(title = "Histogram of Gene Expression for A1BG",
       x = "Expression",
       y = "Frequency") +
  theme_minimal()

```

## Plotting gene expression vs. ferritin (scatterplot)

```{r scatterplot}

```

## Plotting gene expression vs. sex and disease status (boxplot)

```{r boxplot}

```