---
title: "submission1"
author: "Antara Sengupta"
date: "2024-07-27"
output:
  pdf_document: default
  html_document: default
always_allow_html: true
---

## Loading in necessary packages
```{r loading packages}
library(readr)
library(dplyr)
library(ggplot2)
library(tibble)
library(tidyr)
library(dplyr)
```

## Loading Data

```{r loading data}
# loading in the two separate datasets provided to us (metadata & gene exppression data)
series <- read.csv("data/QBS103_GSE157103_series_matrix.csv")
genes <- read.csv("data/QBS103_GSE157103_genes.csv")

# displaying first few rows of each dataframe to get familiarized with the data
head(series)
head(genes)
```
## Formatting and cleaning data

```{r adjusting dataframes}
# goal is to merge both dataframes - we can merge them based on participant ID

# transposing the genes df and turning it back to a df so it can have 'participant_id' as rows (the same rows as the metadata)
genes_transposed <- as.data.frame(t(genes))

# assigning row x with gene name to be the column headers
colnames(genes_transposed) <- as.character(genes_transposed[1, ])

# getting rid of the x row because it has redundant info (same values as column names)
genes_transposed <- genes_transposed[-1, ]

# setting the row name to participant id so we can use this column to merge with metadata
# used this source to find the rownames_to_column function https://forum.posit.co/t/rstudio-rownames-to-column-how-to-find-name-of-first-column-to-use-in-this-function/113079
genes_transposed <- rownames_to_column(genes_transposed, var = "participant_id")

# reshaping the gene so that genes and expression can be their own columns using pivot_longer
# used this source to learn how to implement pivot_longer: https://tidyr.tidyverse.org/reference/pivot_longer.html
genes_long <- genes_transposed %>%
   pivot_longer(
     cols = -participant_id,  #accessing all columns with the exception of participant_id because it is the only column without gene info
     names_to = "gene", #setting name column to gene
     values_to = "expression"  #setting values column to expression
   )

# merging long gene data and series metadata to become one dataframe
all_data <- merge(genes_long, series, by = "participant_id") #merging it on 'participant_id' column

# taking a look at the new comprehensive dataframe
head(all_data)
```


## Choosing focuses of interests: genes, continuous covariates and categorical covariates
gene: AASDHPPT
continuous covariate: ferritin levels
categorical covariate: disease status, gender 
```{r subsetting dataframe}
#subsetting the dataframe so that it only contains columns that I want to perform further analysis on 


#getting rid of all unknown values in the data
#unique(all_data$ferritin.ng.ml.)
#unique(all_data$Age)
#used the above lines to look at unique values, and saw that unknown is formatted as " unknown", so will replace that with NA
all_data[all_data == " unknown"] <- NA

# Drop rows with any NA values
all_data <- na.omit(all_data)

#using select to subset https://www.educative.io/answers/what-is-the-select-function-in-r
covid_data <- all_data %>% select(participant_id, gene,expression, ferritin.ng.ml., sex, disease_status)

#taking a look at the new subsetted data frame
head(covid_data)
```

## Plotting gene expression (histogram)
```{r histogram}
library(tidyverse)
AASDHPPT_data <- covid_data %>%
  filter(gene == "AASDHPPT") %>%
  select(expression) %>%
  mutate(expression = as.numeric(expression))  # Ensure expression is numeric

# Plot histogram
ggplot(AASDHPPT_data, aes(x =as.numeric(expression))) +
  geom_histogram(binwidth = 0.5, color = "black", fill = "blue") +
  labs(title = "Histogram of Gene Expression for AASDHPPT",
       x = "Expression",
       y = "Frequency") +
  theme_minimal() #choosing preferred theme
```

## Plotting gene expression vs. ferritin (scatterplot)

```{r scatterplot}
# referred to these sites when plotting
# https://r-graph-gallery.com/interactive-charts.html
# https://r-graph-gallery.com/scatterplot.html
# http://www.sthda.com/english/wiki/ggplot2-scatter-plots-quick-start-guide-r-software-and-data-visualization

# using plotly to add interactive element to the graph
#install.packages("plotly")
library(plotly)

# subsetting the data so that it only contains rows where gene = AASDHPPT and also the sex variable so it gets rid of the unknown rows
data <- covid_data %>%
  filter(gene == "AASDHPPT")

# certain columns in the dataset with numeric values are stored in character types, setting them to numeric to be able to plot 
data$expression <- as.numeric(data$expression)
data$ferritin.ng.ml. <- as.numeric(data$ferritin.ng.ml. )

scatter_plot <- ggplot(data, aes(x = expression, y = ferritin.ng.ml.)) +
    geom_point(color = "purple",size = 3) + #increasing size of the data points on the graoh
    geom_smooth(method = "lm", se = FALSE, color = "#33FFF7",linetype = "dashed") +
  labs(title = "AASDHPPT Expression and Ferritin Levels Appear Slightly Negatively Correlated",
      subtitle = "Exploring AASDHPPT Expression vs. Ferritin Levels in Human Subjects",
       x= "AASDHPPT Expression" ,
       y = "Ferritin (ng/mL)") +
  theme_bw() # changing to preferred theme +

interactive_scatter <- ggplotly(scatter_plot)

# displaying regular graph and interactive
scatter_plot
interactive_scatter
```

## Plotting gene expression vs.disease status by sex (boxplot)

```{r boxplot}
# creating boxplot
ggplot(data, aes(x = disease_status, y = as.numeric(expression), fill = sex)) +
  geom_boxplot() +
  scale_x_discrete(labels = c("disease state: COVID-19" = "COVID-19 Positive", "disease state: non-COVID-19" = "COVID-19 Negative"))+ # changing the x values to visually cleaner/shorter ones
  labs(title = "Observing patterns of AASDHPPT Expression by Gender and Disease Status", x= "Disease Status" ,y = "AASDHPPT Expression", fill = "Patient Sex" ) + #adding labrls and titles to the graph
  theme_bw() # changing to preferred theme
```

